Statistically-estimated tree composition for the northeastern United States \\ at the time of European settlement

### 2014-12-11 paper.tex
added simon intro material

\usepackage{alltt}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{geometry}
\geometry{verbose,tmargin=1in,bmargin=1in,lmargin=1in,rmargin=1in}
\usepackage{array}
\usepackage{amsmath}
\usepackage{graphicx}

@@

\title{Statistically-estimated tree composition for the northeastern United
States \\
at the time of European settlement}
\maketitle
\begin{quote}
Christopher J. Paciorek, Andrew Thurman, Simon J. Goring, Charlie
V. Cogbill, \\
John W. Williams, David J. Mladenoff, Jody Peters, Jun Zhu, \\
and Jason S. McLachlan {[}author order to be discussed{]}\\
{[}please tell me if you do/don't want your middle initial{]}
\end{quote}
TODO:

@@

\begin{abstract}
We present a data product of the estimated composition of tree taxa
at the time of European settlement of the United States and the statistical
methodology used to produce the product. Composition is defined as
the proportion of stems larger than approximately X cm diameter at
breast height {[}Charlie? Simon? is X roughly 10?{]} in 22 taxonomic
groupings, generally at the genus level. The data come from settlement
survey records that provide raw data that are transcribed and then
aggregated spatially, giving count data. For the western portion of
the domain (Indiana through Minnesota) we use data at survey points
that are aggregated to a regular 8 km grid, while for the eastern
portion (Ohio through Maine), we use data aggregated into townships.
The product is based on a Bayesian statistical model fit to the count
data that estimates composition on a regular 8 km grid. The statistical
model allows us to estimate composition at locations with no data
and to smooth over noise caused by limited counts in locations with
data. Critically, it also allows us to quantify uncertainty in our
composition estimates. We expect this data product to be useful for
understanding the state of vegetation in the northeastern United States
prior to large-scale European settlement. In addition to specific
regional questions, the data product can also serve as a baseline
against which to investigate how forests and ecosystems change after
intensive settlement.
\end{abstract}

@@

{[}insert paragraph (Jack has volunteered) on context and importance
of historical datasets{]}

@@

The raw data are obtained from survey records collated from across
the northeastern U.S. by a number of researchers. For the states of
Minnesota, Wisconsin, Illinois, Indiana, and Michigan (the western
subdomain), data are available at survey point locations and have
been aggregated to a regular 8 km grid in the Albers projection. For
the states of Ohio, Pennsylvania, New Jersey, New York and the six
New England states (the eastern subdomain), data are aggregated at
the township level. There is also data from a single township in Quebec
and a single township in northern Delaware. Data are essentially complete
in Minnesota, Wisconsin and Michigan but data in Illinois and Indiana
represent a sample of the full set of grid cells, with transcription
ongoing. Data for the remaining states are a subset of the full set
of townships. Fig. \ref{fig:domain} shows the domain, indicating
the grid cells and townships with data.

\begin{figure}
\label{fig:domain}

\includegraphics{0_accounts_gen_vis_paciorek_research_jmac_composition_paper_fig1.pdf}

@@

Extensive details on the data are available in \cite{Gori:etal:2015}
and the raw data are available at XXX. The aggregation into taxonomic
groups is primarily at the genus level, but in some cases of monospecific
genera, is at the species level. In some other cases, because of ambiguity
in the common tree names used by surveyors, a group represents trees
from different families (orders?). We model the following taxa: {[}list
all{]}. {[}include description of size cutoff{]}

@@

Our domain is a rectangle covering all of the states, in the Albers
projection, with the rectangle split into 8 km cells, arranged in
a 296 by 180 grid of cells, with the centroid of the cell in the southeast
corner located at (-71000, 58000). For the modeling of the western
subdomain we use the western-most 146 cells. For the modeling of the
western subdomain we use the eastern-most 180 cells and omit 23 cells
in the north and 17 cells in the south. {[}check 23 in N vs 17 in
N; also check what is in netCDF product - should be for the full domain
with missing values?{]}

@@

\begin{table}

@@

Table \ref{tab:score_cell_20percent} shows the results when the proportion
of cells that are held out decreases from 95\% to 80\%, indicating
performance on less sparse data. Table \ref{tab:coverage_cell_fivepercent}
shows performance of the uncertainty estimates. For denser data, unlike
the results when data are more sparse, the SPDE model generally outperforms
the CAR model, but again differences from a practical perspective,
based on mean absolute error, are limited.

\begin{table}
\caption{Predictive ability based on several predictive score criteria for
the CAR and SPDE spatial models when holding out 80\% of entire cells
of data in Minnesota. Smaller values are better, except for coverage,
where values near 0.95 are optimal.}

@@

\begin{table}
\caption{Predictive ability based on several predictive score criteria for
the CAR and SPDE spatial models when holding out 5\% of trees. Smaller
values are better.}

@@

The authors are deeply indebted to all of the researchers over the
years who have preserved, collected, and digitized survey records,
in particular Ed Schools (Michigan DNR, \{ Brugam, ??? - Jody/Charlie/David/Jason/Simon
please add names here\}. This work was carried out by the PalEON Project
with support from the National Science Foundation MacroSystems Program
through grants EF-1065656, EF-1241868, DEB-1241874 and DEB-1241868.

@@

\bibliography{/accounts/gen/vis/paciorek/bibfiles/abbrev.stat,/accounts/gen/vis/paciorek/bibfiles/abbrev.other,/accounts/gen/vis/paciorek/bibfiles/spatstat,/accounts/gen/vis/paciorek/bibfiles/statgeneral,/accounts/gen/vis/paciorek/bibfiles/paciorek,/accounts/gen/vis/paciorek/bibfiles/ecology}

@@

The conditional distribution for $W_{mjp}$ given the other unknowns
in the model and the data is as follows. Let $\mbox{TN}(a,b,\mu,\tau^{2})$
denote the truncated normal distribution with mean parameter $\mu$
and variance parameter $\tau^{2}$, truncated below by $a$ and above
by $b$.

>>>>

\usepackage{alltt}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{geometry}
\geometry{verbose,tmargin=1in,bmargin=1in,lmargin=1in,rmargin=1in}
\usepackage{array}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage[authoryear]{natbib}

@@

\title{Statistically-estimated tree composition for the northeastern United
States \\
at the time of European settlement}
\maketitle
\begin{quote}
Christopher J. Paciorek, Andrew Thurman, Simon J. Goring, Charlie
V. Cogbill, \\
John W. Williams, David J. Mladenoff, Jody A. Peters, Jun Zhu, \\
and Jason S. McLachlan {[}author order to be discussed{]}\\
{[}please tell me if you do/don't want your middle initial{]}
\end{quote}
TODO:

@@

%
\begin{abstract}
We present a data product of the estimated composition of tree taxa
at the time of European settlement of the United States and the statistical
methodology used to produce the product. Composition is defined as
the proportion of stems larger than approximately X cm diameter at
breast height {[}Charlie? Simon? is X roughly 10?{]} in 22 taxonomic
groupings, generally at the genus level. The data come from settlement
survey records that provide raw data that are transcribed and then
aggregated spatially, giving count data. The domain is divided into
two regions, western (Indiana through Minnesota) and eastern (Ohio
to Maine). Public Land Survey point data in the western region is
aggregated to a regular 8 km grid, while data in the eastern region,
from Town Proprietor Surveys is aggregated at the township level.
The product is based on a Bayesian statistical model fit to the count
data that estimates composition on a regular 8 km grid. The statistical
model allows us to estimate composition at locations with no data
and to smooth over noise caused by limited counts in locations with
data. Critically, it also allows us to quantify uncertainty in our
composition estimates. We expect this data product to be useful for
understanding the state of vegetation in the northeastern United States
prior to large-scale European settlement. In addition to specific
regional questions, the data product can also serve as a baseline
against which to investigate how forests and ecosystems change after
intensive settlement.
\end{abstract}

@@

Historical datasets provide critical context to understand forest
ecology. Historical datasets allow researchers to define 'baseline'
conditions in conservation management, to understand ecosystem processes
at decadal and centennial scales, and, particularly in regions with
widespread land use change, to understand the extent to which forest
conversion and regeneration differ from the original forest cover.

Euro-American settlement and subsequent land use change occured in
a time transient fashion across North America, and surveys of vegetation
followed a similar pattern. Early surveys (from 1620 until 1825) in
the northeastern United States were areal surveys at a township level
\citep{cogbill2002forests,thompson2013four}, which we call the Town
Proprietor Survey (TPS). These surveys reporting percent composition
of major tree species using sometimes inconsistent common names. Later
surveys, after the establishment of the Public Land Survey System
(PLSS) - from 1819 to 1904 - were point estimates along a regular
grid, with 1 mile spacing \citep{schulte2001original,bourdo1956review,goring2015composition}.
Survey instructions during the PLSS varied through time, and by point
type. This required the application of spatially varying correction
factors \citep{cogbill2015corrections,goring2015composition} to accurately
assess stem density, basal area and biomass from the early settlement
records.

Logging, agriculture and abandonment has left an indelible mark on
forests in the northeastern United States \citep{foster1998land,goring2015composition,rhemtulla2009legacies,thompson2013four},
however most studies have studied these effects at the state level
or below \citep{rhemtulla2009historical,friedman2005regional}, and
with varying spatial resolution, from townships (36 square miles)
to forest zones of hundreds or thousands of square miles. \citet{goring2015composition}
provide a dataset across the upper Midwest that is resolved to an
8 x 8 km grid size, providing broad spatial coverage at a meso-spatial
scale that can be compared to the Forest Inventory and Analysis products
\citet{gray2012forest}. The limitation of this dataset is that it
provides only estimates of variance within cell for PLSS data in the
western region, and no township estimates of error for regions surveyed
as part of the Town Proprietor Survey.

Properly assessing uncertainty in ecological data is imperative to
understanding and modelling ecological processes \citep{cressie2009accounting}.
In this way, a model that can account for the spatial structure of
the underlying PLSS and TPS data, and provide reliable estimates of
uncertainty across the northeastern United States provides an enormously
valuable tool for researchers interested in the ecological structure
and function of forests at longer time scales.

@@

The raw data are obtained from survey records collated from across
the northeastern U.S. by a number of researchers. For the states of
Minnesota, Wisconsin, Illinois, Indiana, and Michigan (the western
subdomain), data are available at survey point locations and have
been aggregated to a regular 8 km grid in the Albers projection. For
the states of Ohio, Pennsylvania, New Jersey, New York and the six
New England states (the eastern subdomain), data are aggregated at
the township level. There is also data from a single township in Quebec
and a single township in northern Delaware. Data are essentially complete
in Minnesota, Wisconsin and Michigan but data in Illinois and Indiana
represent a sample of the full set of grid cells, with survey record
transcription ongoing. Data for the remaining states are a subset
of the full set of townships. Fig. \ref{fig:domain} {[}Fig. 1 - Lyx
is having a numbering problem{]} shows the domain, indicating the
grid cells and townships with data.

\begin{figure}
\label{fig:domain}\includegraphics{0_accounts_gen_vis_paciorek_research_jmac_composition_paper_fig1.pdf}

@@

Extensive details on the data are available in \cite{Gori:etal:2015}
and the raw data are available at XXX. The aggregation into taxonomic
groups is primarily at the genus level, but in some cases of monospecific
genera, is at the species level. We model the following 22 taxa plus
an ``other hardwood'' category: Atlantic white cedar (Chamaecyparis
thyoides), Ash (Fraxinus spp.), Basswood (Tilia americana), Beech
(Fagus grandifolia), Birch (Betula spp.), Black gum/sweet gum (Nyssa
sylvatica and Liquidambar styraciflua), Cedar/juniper (what is in
here other than Juniperus and is the Juniperus only Juniperus virginiana),
Cherry (Prunus spp.), Chestnut (Castanea dentata), Dogwood (Cornus
spp.), Elm (Ulmus spp. or Ulmus americana???) , Fir (Abies spp.),
Hemlock (Tsuga canadensis), Hickory (Carya spp.), Ironwood (Carpinus
caroliniana and Ostrya virginiana (anything else?)), Maple (Acer spp.),
Oak (Quercus spp.), Pine (Pinus spp.), Poplar/tulip poplar (Populus
spp. and Liriodendron tulipifera), Spruce (Picea spp.), Tamarack (Larix
laricina), Walnut (Juglans nigra). {[}Simon/Charlie to check my Latin
and whether I've left out anything that falls into a given common
name{]} Note that in several cases (e.g., black gum/sweet gum, ironwood,
poplar/tulip poplar, cedar/juniper), because of ambiguity in the common
tree names used by surveyors, a group represents trees from different
families and even orders. For the western subdomain we do not fit
statistical models for Atlantic white cedar and chestnut as these
have 0 and 5 trees present in the dataset, respectively.

{[}include description of size cutoff{]} {[}Simon, Jody mentioned
that ther are 402 trees in IN below \textasciitilde{}10 cm and at
least 43 in IL. Do you screen those out or do we just live with it?{]}

@@

Our domain is a rectangle covering all of the states, in the Albers
projection (NAD 1983 Great Lakes and St. Lawrence Albers) {[}Simon,
please confirm this language is right{]}, with the rectangle split
into 8 km cells, arranged in a 296 by 180 grid of cells, with the
centroid of the cell in the southeast corner located at (-71000, 58000).
For the modeling of the western subdomain we use the western-most
146 by 180 grid of cells. For the modeling of the eastern subdomain
we use the eastern-most 180 by 180 grid of cells and then omit 23
cells in the north and 17 cells in the south. {[}check 23 in N vs
17 in N; also check what is in netCDF product - should be for the
full domain with missing values?{]}

@@

%
\begin{table}

@@

Table \ref{tab:score_cell_20percent} shows the results when the proportion
of cells that are held out decreases from 95\% to 80\%, indicating
performance on less sparse data. Table \ref{tab:coverage_cell_20percent}
shows performance of the uncertainty estimates. For denser data, unlike
the results when data are more sparse, the SPDE model generally outperforms
the CAR model, but again differences from a practical perspective,
based on mean absolute error, are limited.

%
\begin{table}
\caption{Predictive ability based on several predictive score criteria for
the CAR and SPDE spatial models when holding out 80\% of entire cells
of data in Minnesota. Smaller values are better, except for coverage,
where values near 0.95 are optimal.}

@@

%
\begin{table}
\caption{Predictive ability based on several predictive score criteria for
the CAR and SPDE spatial models when holding out 5\% of trees. Smaller
values are better.}

@@

The authors are deeply indebted to all of the researchers over the
years who have preserved, collected, and digitized survey records,
in particular Ed Schools (Michigan DNR), \{ Brugam, ??? - Jody/Charlie/David/Jason/Simon
please add names here\}. This work was carried out by the PalEON Project
with support from the National Science Foundation MacroSystems Program
through grants EF-1065656, EF-1241868, DEB-1241874 and DEB-1241868.

@@

\bibliography{/accounts/gen/vis/paciorek/bibfiles/abbrev.stat,/accounts/gen/vis/paciorek/bibfiles/abbrev.other,/accounts/gen/vis/paciorek/bibfiles/spatstat,/accounts/gen/vis/paciorek/bibfiles/statgeneral,/accounts/gen/vis/paciorek/bibfiles/paciorek,/accounts/gen/vis/paciorek/bibfiles/ecology,goring_intro}

@@

The conditional distribution for $W_{ijp}$ given the other unknowns
in the model and the data is as follows. Let $\mbox{TN}(a,b,\mu,\tau^{2})$
denote the truncated normal distribution with mean parameter $\mu$
and variance parameter $\tau^{2}$, truncated below by $a$ and above
by $b$.
